---
title: "Exploratory Analysis"
author: "Joe Watson"
date: "07/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Analysis

Crucial to all good statistical analyses is a thorough exploratory analysis. In this document, we will introduce a few tecniques for developing an understanding of our dataset, including an understanding of its limitations.

Note that all the workshop scripts and files can be found at <http://rmarkdown.rstudio.com> and <...> respectively.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
